---
title: "Quanta versus Funda Mental"
description: |
  A quick introduction to what mean by quantamental.
author:
  - name: Jonathan K. Regenstein
    url: https://fintech.gatech.edu/#/people
  - name: Sudheer Chava
    url: https://fintech.gatech.edu/#/people
date: 08-18-2020
preview: q-f.png
base_url: http://quantamentalr.com/
categories:
  - quantamental
twitter:
  site: "@jkregenstein"
  creator: "@jkregenstein"
output: 
  distill::distill_article:
    toc: true
    toc_depth: 3
    self_contained: false
---

```{r setup, include=FALSE}

library(tidyverse)
library(tidyquant)
library(highcharter)
library(timetk)
library(tibbletime)
library(scales)
library(httr)
library(jsonlite)
library(plotly)
library(here)
library(riingo)
library(reactable)

api_key <-  config::get("tiingo", file = "config.yml")

knitr::opts_chunk$set(message=FALSE, 
                      warning=FALSE,
                      comment = NA, 
                      echo = TRUE)
```

What is quanta versus funda mental analysis? 

In FM, items from financial statements are inputs to try to estimate a valuation. Almost universally, this involves updating values in a spreadsheet with a complicated link structure that ultimately generates a value for a company, which leads to a price target for an equity. If that price targe is above, the current price, the equity will usually be rated a 'buy', and so on. In QM, financial statement items can be treated in two ways. First, as features, or the basis for engineered features, that are used to model and predict the movement of an equity's price. In other words, in FM, items are numeric inputs, that are entered into equations, and those equations generate a number. In QM, items are features that generate model results, which are used to estimate future outcomes. 

Relatedly, in QM, those financial statement items can themselves be the the variables of interest. We might be attempting to forecast or nowcast financial statement items, in order to forecast or nowcast price movements. 

In FM, the relationship between financial statement items and the variable of interest, usually price, is not statistical, it's the not the result of estimating model parameters. It's the result of assumptions or calculations about the relationship between items and price. In QM, the relationships can be modeled or estimated using statistical learning or machine learning, which depends on larger data sets and generally longer history than FM.

### Definitions

Let's first review the definitions for each of the data codes. We are accessing this data via an API and very often an API will supply a data dictionary similar to these definitions.

```{r}
definitions <-
  httr::GET(
    url = str_glue("https://api.tiingo.com/tiingo/fundamentals/definitions?format=csv&token={api_key}")
    ) %>%
  content(as = "parsed")  %>%
  select(statementType, everything()) %>%
  arrange(statementType)  

definitions %>% 
  head() %>% 
  reactable()
```




### Import Financials from Tiingo


```{r, echo =  FALSE}
aapl_financial_statements <- 
  GET(
  url = str_glue(
    "https://api.tiingo.com/tiingo/fundamentals/aapl/statements?format=csv&startDate=2018-01-01&endDate=2020-01-31&token={api_key}"
  )) %>%
  content(as = "parsed")  %>% 
  left_join(definitions)
```

```{r,eval=FALSE}
aapl_financial_statements <- 
  GET(
  url = str_glue(
    "https://api.tiingo.com/tiingo/fundamentals/aapl/statements?format=csv&
    startDate=2018-01-01&endDate=2020-01-31&token={api_key}"
  )) %>%
  content(as = "parsed")  %>% 
  left_join(definitions)
```

We just imported all of the financial statements for Apple from 2018 to beginning of 2020. Let's `filter()` down to just the 10K Income Statement for 2019. We get the 10K by setting `filter(quarter == 0)`, which indicates we want the annual report.


```{r}
aapl_2019_annual_is <-  
  aapl_financial_statements %>% 
  filter(statementType == "incomeStatement",
         year == 2019,
         quarter == 0)
```

Tiingo does not include an entry for Cost of Goods Sold, or COGS. That's a common item and we can calculate it from the difference between `revenue` and `grossProfit` items. We built a custom function to do this called `cogs_get_funs()` that we can pass an income statement to in order to get back the COGS. 

To learn about how we constructed this function, have a look at [this post](https://quantamental-r-draft.netlify.app/posts/2020-08-19-getting-functional-for-cogs/).


```{r, echo = FALSE}
cogs_get_funs <- function(income_statement){
  income_statement %>% 
    filter(dataCode %in% c("revenue", "grossProfit")) %>% 
    group_by(date) %>% 
    select(date, dataCode, value) %>% 
    pivot_wider(names_from = "dataCode", values_from = "value") %>% 
    summarise(value = revenue - grossProfit) %>% 
    mutate(dataCode = "cogs", 
           name = "Cost of Goods Sold",
           units = "$",
           statementType = "incomeStatement")
}
```


```{r}
cogs_get_funs(aapl_2019_annual_is)
```

### Visualize Income Statement

```{r}
aapl_2019_annual_is %>% 
  # select(-year, -quarter) %>% 
  bind_rows(cogs_get_funs(.)) %>% 
  filter(
    dataCode %in% c("revenue", "cogs", "grossProfit", "rnd", "sga", "opinc", "intexp", "ebt", "taxExp", "netinc"),
    date == max(date)
  ) %>% 
  arrange(value) %>% 
  mutate(
    value = case_when(dataCode %in% c("cogs", "taxExp", "intexp", "rnd", "sga") ~ value * -1,
                      TRUE ~ value),
    measure = case_when(dataCode %in% c("grossProfit", "opinc", "ebt", "netinc") ~ "total",
                        TRUE ~ "relative"),
    dataCode = fct_relevel(
      dataCode,
      c("revenue",
        "cogs",
        "grossProfit",
        "rnd",
        "sga",
        "opinc",
        "intexp",
        "ebt",
        "taxExp",
        "netinc"
        )
    ),
    name = fct_relevel(
      name,
      c("Revenue",
        "Cost of Goods Sold",
        "Gross Profit",
        "Research and Development",
        "Selling, General and Administrative",
        "Operating Income",
        "Interest Expence",
        "Earnings before tax",
        "Tax Expense",
        "Net Income"
        )
    )
  ) %>% 
  arrange(dataCode) %>% 
  select(dataCode, value, measure, name) %>% 
  plot_ly(name = "Income Statement", type = "waterfall", measure = ~measure, 
  x = ~dataCode, textposition = "outside", y= ~value,
  text = ~str_glue("{name}
                   ${scales::label_number_si(accuracy = 0.1)(value)}"),
  hoverinfo = 'text',
  showlegend = F
  ) %>%
  # connector = list(line = list(color= "rgb(63, 63, 63)"))) %>%
  layout(title = str_glue("Income statement 2019"),
        xaxis = list(title = ""),
        yaxis = list(title = "",
                     tickprefix = "$"),
        autosize = TRUE,
        showlegend = TRUE)
```



We start with Revenue, and subtract COGS to get Gross Profit. We then subtract Research and Development Costs, and Selling, General and Adminstrative Costs to get Operating Income (or EBIT). From there, we move to EBT and finally Net Income. 

Similar to what we did with the [COGS function](link to url), we could wrap this entire process into a function that returns a chart, instead of returning a number or a data frame. 

```{r}
income_statement_waterfall_charter <- function(income_statement){
  income_statement %>% 
  # select(-year, -quarter) %>% 
  bind_rows(cogs_get_funs(.)) %>% 
  filter(
    dataCode %in% c("revenue", "cogs", "grossProfit", "rnd", "sga", "opinc", "intexp", "ebt", "taxExp", "netinc"),
    date == max(date)
  ) %>% 
  arrange(value) %>% 
  mutate(
    value = case_when(dataCode %in% c("cogs", "taxExp", "intexp", "rnd", "sga") ~ value * -1,
                      TRUE ~ value),
    measure = case_when(dataCode %in% c("grossProfit", "opinc", "ebt", "netinc") ~ "total",
                        TRUE ~ "relative"),
    dataCode = fct_relevel(
      dataCode,
      c("revenue", 
        "cogs",
        "grossProfit",
        "rnd",
        "sga",
        "opinc",
        "intexp",
        "ebt",
        "taxExp",
        "netinc"
        )
    ),
    name = fct_relevel(
      c("Revenue",
        "Cost of Goods Sold",
        "Gross Profit",
        "Research and Development",
        "Selling, General and Administrative",
        "Operating Income",
        "Interest Expence",
        "Earnings before tax",
        "Tax Expense",
        "Net Income"
        )
    )
  ) %>% 
  arrange(dataCode) %>% 
  select(dataCode, value, measure, name) %>% 
  plot_ly(name = "Income Statement", type = "waterfall", measure = ~measure,
  x = ~dataCode, textposition = "outside", y = ~value, 
  text = ~str_glue("{name}
                   ${scales::label_number_si(accuracy = 0.1)(value)}"),
  hoverinfo = 'text',
  showlegend = F
  ) %>% 
  # connector = list(line = list(color= "rgb(63, 63, 63)"))) %>%
  layout(title = str_glue("Income statement 2019"),
        xaxis = list(title = "", tickfont = "16", ticks = "outside"),
        yaxis = list(title = "",
                     tickprefix = "$"),
        autosize = TRUE,
        showlegend = TRUE)
}
```


Now we pass the `aapl_2019_annual_is` data frame to the function. 

```{r}
income_statement_waterfall_charter(aapl_2019_annual_is)
```

Note that we actually called our `cogs_get_funs()` function inside that visualizer function. We can call functions inside of other functions. 

